\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-12-15 doc/paper-jss/sdmTMB-paper.Rnw
Add linenumbers + response input

@@

    Eric J. Ward\\
    Northwest Fisheries Science Center\\
    National Oceanic and Atmospheric Administration, National Marine
    Fisheries Service, Seattle, WA, USA\\

    Philina A. English\\
    Pacific Biological Station\\
    Fisheries and Oceans Canada, Nanaimo, BC, Canada\\

    Lewis A. K. Barnett\\
    Alaska Fisheries Science Center\\
    National Marine Fisheries Service, National Oceanic and
    Atmospheric Administration, Seattle, WA, USA\\

    James T. Thorson\\
    Alaska Fisheries Science Center\\
    National Marine Fisheries Service, National Oceanic and
    Atmospheric Administration, Seattle, WA, USA
  }

@@

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

@@

In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
One such approach is the SPDE approach.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develop the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} summarizes the approach and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provide a recent review of the approach and its applications over the last decade.

For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
% A19
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.

@@

The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field.
This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
We then use the function \fct{spread\_sims} to draw 200 simulations for the parameters themselves.
Because the simulations are stored in different dimensions, the random field draws must be transposed \code{t()} before combining the vector of main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
For this example, we use thresholds of 0.1 and 0.9 representing CIs of 80\%.

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
We think users familiar with \fct{stats::glm}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
Users familiar with \INLA\ will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \sdmTMB\ more straightforward.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
% TODO: mention this in response doc

@@

While our examples focused on applications to ecological data, the SPDE approach and the functionality of \sdmTMB\ has applications in many other fields.
Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.

@@

<<inlabrufit-timing, echo=TRUE, cache=TRUE>>=
# convert to sp SpatialPointsDataFrame first:
dat_sp <- sp::SpatialPointsDataFrame(
  cbind(sim_dat$X, sim_dat$Y),
  proj4string = sp::CRS(
    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
+ +y_0=0 +datum=NAD83 +units=km +no_defs"
  ), data = sim_dat
)
components <- observed ~ -1 + Intercept(1) + a1 +
  spatrf(main = coordinates, model = spde)
like <- like(observed ~ Intercept + a1 + spatrf,
  family = "gaussian", data = dat_sp
)
fit_bru <- bru(
  like,
  components = components,
  options = bru_options(
    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
    bru_max_iter = 1, num.threads = "1:1"
  )
)
@

@@

>>>>

% TODO START FOR REVIEW ONLY
\usepackage{lineno}
\definecolor{darkgrey}{HTML}{A9A9A9}
\renewcommand\linenumberfont{\normalfont\bfseries\small\color{darkgrey}}
\modulolinenumbers[2]
\usepackage[dvipsnames]{xcolor}
\definecolor{niceblue}{HTML}{236899}
\newcommand{\rev}[1]{{\color{niceblue} #1}}
\newcommand{\Rev}[1]{\label{#1}\linelabel{#1}}
\newcommand{\lr}[1]{line~\lineref{#1}}
% TODO END FOR REVIEW ONLY

@@

    Eric J. Ward\\
    Northwest Fisheries Science Center\\
    National Oceanic and Atmospheric Administration, National Marine
    Fisheries Service, Seattle, WA, USA\\
    E-mail: \email{eric.ward@noaa.gov}\\

    Philina A. English\\
    Pacific Biological Station\\
    Fisheries and Oceans Canada, Nanaimo, BC, Canada\\
    E-mail: \email{philina.english@dfo-mpo.gc.ca}\\

    Lewis A. K. Barnett\\
    Alaska Fisheries Science Center\\
    National Marine Fisheries Service, National Oceanic and
    Atmospheric Administration, Seattle, WA, USA\\
    E-mail: \email{lewis.barnett@noaa.gov}\\

    James T. Thorson\\
    Alaska Fisheries Science Center\\
    National Marine Fisheries Service, National Oceanic and
    Atmospheric Administration, Seattle, WA, USA\\
    E-mail: \email{james.thorson@noaa.gov}\\
  }

@@

\linenumbers

@@

Here, we introduce the \R\ package \sdmTMB, which implements geostatistical spatial and spatiotemporal GLMMs using \TMB\ for model fitting and \pkg{fmesher} or \INLA\ to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} \citep{wood2017a} for a specific class of spatial and spatiotemporal models.
\Rev{B2}Many features of \sdmTMB\ may be found in other software (Table~\ref{tab:functionality}), but to date no package has included all in a single package.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \TMB: \sdmTMB), although the package is widely applicable to other fields and any geostatistical data collected continuously in space and discretely in time.

@@

\Rev{B4a}In practice, working with the dense covariance matrix $\bm{\Sigma}$ is computationally expensive
and methods for working directly with its inverse, the precision matrix $\bm{Q}$, are more efficient ($\bm{Q} = \bm{\Sigma}^{-1}$) \citep{simpson2012}.
One such approach is the SPDE approach.
A full description of the SPDE approach is beyond the scope of this paper.
Instead, we refer to the following literature:
\citet{lindgren2011} develop the approach.
\citet{lindgren2015} and \cite{bakka2018review} summarize the SPDE approach for spatial modelling in the context of \INLA.
The second chapter of \citet{krainski2018} provides an overview of the SPDE approach to spatial modelling with a focus on linking theory to code.
\citet{miller2019} summarizes the approach and illustrates its equivalence to penalized smoothing approaches.
\citet{lindgren2022ten} provide a recent review of the approach and its applications over the last decade.

\Rev{B4b}For a user of \sdmTMB, the following are the important elements to understand.
First, the SPDE approach links Gaussian random fields (GRFs) that have a Mat\'ern covariance function to Gaussian \emph{Markov} random fields (GMRFs) in such a way that a GMRF can be a good approximation to a GRF \citep{lindgren2011}.
This means that GRF models can be computationally approximated as GMRFs.
By working with GMRFs, one can take advantage of theory developed to estimate their sparse precision matrix efficiently \citep{rue2005gmrf, lindgren2011} and avoid the inversion of large dense matrices.
Second, the SPDE involves piece-wise linear basis functions that are defined by a triangulation over the spatial area of interest \citep{lindgren2011}.
Commonly, this is referred to as a ``mesh''.
The properties of this mesh (e.g., its resolution and how far it extends beyond the data) affect the accuracy and computational efficiency of the SPDE approach \citep{lindgren2011}.
Third, working with the SPDE approach and the precision matrix of a GMRF ($\bm{Q}$) introduces an alternative parameter $\tau$, which scales the precision matrix, and involves three sparse matrices associated with the mesh ($\bm{C}$, $\bm{G}_1$, and $\bm{G}_2$) \citep{lindgren2011},
\[
\bm{Q} = \tau^2 \left(\kappa^4\bm{C} + 2\kappa^2\bm{G}_1 + \bm{G}_2 \right),
\]
where $\kappa$ is the Mat\'ern decorrelation rate as before.
We can calculate the marginal variance of the Mat\'ern random field as $\sigma^2 = \left (4 \pi \tau^2 \kappa^2 \right)^{-1}$.

@@

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{statsv::optimHess} \citep{gay1990, r2021}.
\Rev{A19}Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
bias-corrected versions of derived quantities based on non-linear transformations of random effects are possible \citep{thorson2016bias} (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

\sdmTMB\ models can include penalized likelihoods by assigning priors (penalties) to model parameters through the \fct{sdmTMB} \code{prior} argument (\code{?sdmTMBpriors}).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \TMB\ \citep{breivik2021, osgood-zimmerman2023}, penalized complexity (PC) priors \citep{simpson2017, fuglstad2019} (\code{?pc\_matern}) can constrain the spatial range and variance parameters.
\Rev{A4}These priors or penalties are available both with maximum likelihood estimation and with MCMC sampling.
If a model will be passed to \pkg{tmbstan} with priors, a \code{bayesian} logical argument should be set to \code{TRUE} to enable Jacobian adjustments for changes of variables (priors applied to parameters that are internally transformed) \citep{carpenter2017}.

@@

We can generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from the model on a grid covering the area of interest and summing the predicted biomass with the \fct{get\_index} function (Figure \ref{fig:dog-index-plot}).
\Rev{A19b}We supply the grid cell area (4 km\textsuperscript{2}) to the \code{area} argument and specify \code{bias\_correct = TRUE} to enable a bias correction needed due to the non-linear transformation of the random effects \citep{thorson2016bias}.

@@

\Rev{B7}The spatially varying effect at any point in space is the combination of the main effect and SVC random effect for \code{nao}.
Mean estimates of the SVC random effect are available in the output of \fct{predict.sdmTMB} in a column starting with \code{zeta\_s} (in this case, \code{zeta\_s\_nao}).
However, we might wish to combine the fixed and random components of a spatially varying effect and assess the uncertainty of these combined predictions.
We illustrate a way of accomplishing this by simulating from the parameter precision matrix assuming the joint parameter covariance matrix is multivariate normal.
We do this by specifying a non-null number of simulation draws to \code{nsim} in \fct{predict.sdmTMB}.
By default, \code{nsim > 0} will return a matrix of draws for the overall prediction.
Here, we instead specify that we want to return draws for the \code{zeta\_s} ($\zeta_{\bm{s}}$) random field, which is the SVC random field.
This returns a matrix where each row matches a row of \code{newdata} and each column is a simulation draw.
We then use the function \fct{spread\_sims} to draw 200 simulations for the parameters themselves.
Because the simulations are stored in different dimensions, the random field draws must be transposed \code{t()} before combining the vector of main effect draws (\code{sims\$nao}) with the random field values \code{zeta\_s}.
Next, we can calculate the median, and upper and lower quantiles for each column of data, which correspond to the rows in the data provided.
For this example, we use thresholds of 0.1 and 0.9 representing CIs of 80\%.

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table~\ref{tab:functionality}), we suggest the primary differences are the user interface and speed.
We think users familiar with \fct{stats::glm}, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \sdmTMB\ most approachable.
Users familiar with \INLA\ will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code \citep{miller2019} and \INLA/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models; alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting ``simple'' univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \sdmTMB\ more straightforward.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \sdmTMB\ or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
\Rev{B3}Overall, we think there is value in that \sdmTMB\ unites functionality useful in many applied settings into a single package.
% TODO: mention this in response doc

@@

While our examples focused on applications to ecological data, the SPDE approach and the functionality of \sdmTMB\ has applications in many other fields.
\Rev{B9a}Examples include spatial models of disease spread \citep{moraga2021}, spatial econometric models of quantities such as housing prices \citep{bivand2014}, analyzing medical imaging data such as MRI scans \citep{naseri2022}, and geophysical models of seismic waves following earthquakes \citep{zhang2015}.
\Rev{B9b}The \sdmTMB\ model is further relevant to what is commonly referred to as spatial \citep{elhorst2010, lee2010} and dynamic spatial panel data models \citep{elhorst2012} in econometrics.
These examples underscore the versatile nature of the SPDE approach and the potential uses of \sdmTMB\ across various scientific and industrial sectors.

@@

% weird eval errors as of 2023-12-14 when in Rnw, hence eval=FALSE

<<inlabrufit-timing, echo=TRUE, cache=TRUE, warning=FALSE, eval=FALSE>>=
dat_sp <- sp::SpatialPointsDataFrame(
  cbind(sim_dat$X, sim_dat$Y),
  proj4string = sp::CRS(
    "+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000
+ +y_0=0 +datum=NAD83 +units=km +no_defs"
  ), data = sim_dat
)
components <- observed ~ -1 + Intercept(1) + a1 +
  spatrf(main = coordinates, model = spde)
like <- like(observed ~ Intercept + a1 + spatrf,
  family = "gaussian", data = dat_sp
)
fit_bru <- bru(
  like,
  components = components,
  options = bru_options(
    control.inla = list(int.strategy = "eb", strategy = "gaussian"),
    bru_max_iter = 1, num.threads = "1:1"
  )
)
@

@@

% \input{response}
